---
title: "EDA"
author: "Malavika Mampally"
date: "2024-12-08"
output: html_document
---

```{r}
library(tidyverse)
```

**I. Load the 3 data sets:**

1.  Pairwise Correlation of FC between regions

    Dimensions: 623 x 2278

    Contains FC correlations for 2278 pairs of brain regions of 623 subjects

2.  Subj_175_Traits

    Dimensions: 623 x 123

    Contains 123 Subj_175_Traits of 623 subjects

3.  Confounding Variables

    Dimensions: 623 x 2

    Contains age and sex of 623 subjects

```{r}
##code
data=cbind(Subj_confound_var,Subj_175_Traits,Pairwise_Corr_FC)
```

**II. Create a population plot to visualize demographics**

```{r}
data1 <- data %>%
  mutate(Age = cut(Age, breaks = c(0, 20, 23, 26, 29, 32, 35, 38, Inf), 
                         labels = c("0-20", "20-23", "23-26", "26-29", "29-32", "32-35", 
                                    "35-38", "38+"), right = FALSE))
# Convert Gender to a categorical variable
summary_data <- data1 %>%
  group_by(Age, Gender) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(
    Gender = factor(Gender, levels = c(1, 2), labels = c("Female", "Male")),  # Convert Gender to categorical
    count = ifelse(Gender == "Male", -count, count)  # Negative counts for Male
  )

# Plot population pyramid
population->ggplot(summary_data, aes(x = Age, y = count, fill = Gender)) +
  geom_bar(stat = "identity", width = 0.8) + geom_text(aes(label = abs(count)),  # Add labels for counts
                                                       position = position_stack(vjust = 0.5),  # Position text in the middle of the bar
                                                       color = "black", size = 3.5)+
  coord_flip() +
  scale_y_continuous(labels = abs) +  # Ensure y-axis labels are positive
  scale_fill_manual(values = c("Male" = "lightblue", "Female" = "royalblue")) +
  labs(
    title = "Population Pyramid",
    x = "Age Groups (Years)",
    y = "Count",
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "top"
  )
##ggsave("population.png", plot = population)

```

**III. Correlation**

```{r}
library(corrplot)
subset_data=data[,c(1:7,77:89,126:129,2399:2403)]
correlation1=corrplot(cor(subset_data))
##ggsave("correlation1.png", plot = correlation1)

```

**IV. Confounding Variables correlation tests**

```{r}

##Brain Region v/s Age Correlation test

corr_brain_age <- apply(Pairwise_Corr_FC, 2, function(br) cor(br, data$Age, use = "complete.obs"))

#store the p value from the tests for each pair
pvals_brain_age <- apply(Pairwise_Corr_FC, 2, function(br) cor.test(br, data$Age)$p.value)

##Brain Region v/s Gender ANOVA Test

pvals_brain_gender <- apply(Pairwise_Corr_FC, 2, function(br) {
  anova_result <- aov(br ~ data$Gender)
  summary(anova_result)[[1]]["data$Gender", "Pr(>F)"]
})

##Traits v/s Age Correlation Test

corr_traits_age <- apply(Subj_175_Traits, 2, function(trait) cor(trait, data$Age, use = "complete.obs"))

#store the p value from the tests for each pair
pvals_traits_age <- apply(Subj_175_Traits, 2, function(trait) cor.test(trait, data$Age)$p.value)

##Traits v/s Gender ANOVA Test

pvals_traits_gender <- apply(Subj_175_Traits, 2, function(trait) {
  anova_result <- aov(trait ~ data$Gender)
  summary(anova_result)[[1]]["data$Gender", "Pr(>F)"]
})

#Threshold for significance
alpha <- 0.05

#For brain regions
significant_brain_age <- sum(pvals_brain_age < alpha) #counts instances where p value is less than 0.05
significant_brain_gender <- sum(pvals_brain_gender < alpha)

# For Traits
significant_traits_age <- sum(pvals_traits_age < alpha,na.rm = T)
significant_traits_gender <- sum(pvals_traits_gender < alpha)

```

**V. Plot to show correlation with Age**

```{r}
#find index of significant brain regions

significant_brain_age <- which(pvals_brain_age < alpha)  #for age
significant_brain_gender <- which(pvals_brain_gender < alpha)  #for gender

# Extract the first 5 significant variables for plotting
sig_brain_age_vars <- Pairwise_Corr_FC[, significant_brain_age[1:5]]
sig_brain_gender_vars <- Pairwise_Corr_FC[, significant_brain_gender[1:5]]

# Similarly find index of significant traits
significant_traits_age <- which(pvals_traits_age < alpha)
significant_traits_gender <- which(pvals_traits_gender < alpha)

# Extract the first 5 significant traits for plotting
sig_traits_age_vars <- Subj_175_Traits[, significant_traits_age[1:5]]
sig_traits_gender_vars <- Subj_175_Traits[, significant_traits_gender[1:5]]

#Plot mean  correlation among different age groups

#Create age bins
age_bins <- cut(data$Age, breaks = seq(20, 40, by = 3), include.lowest = TRUE)

# Calculate mean values for each bin
mean_values <- tapply(sig_brain_age_vars[, 1], age_bins, mean)
# Plot with custom x-axis labels
plot_regions_age=plot(seq(22, 38, by = 3), mean_values, type = "b", pch = 16,
     xlab = "Age Group", ylab = "Mean Correlation",col="red",
     main = "Mean Brain Region Correlation vs. Age Group", xaxt = "n")
+
axis(1, at = seq(22, 38, by = 3), labels = levels(age_bins))


##ggsave("brain_age.png,plot=plot_regions)
```

```{r}
mean_values_tr=tapply(sig_traits_age_vars[,2],age_bins,mean)
# Plot with custom x-axis labels
plot_traits_age=plot(seq(22, 38, by = 3), mean_values_tr, type = "b", pch = 16,
     xlab = "Age Group", ylab = "Mean Reaction time on a memory test",col="blue",
     main = "Trait #9 vs. Age Group", xaxt = "n")+
axis(1, at = seq(22, 38, by = 3), labels = levels(age_bins))

##ggsave("traits_age.png",plot=plot_traits_age)
```

**VI. Plot to show correlation with Gender**

```{r}
male_data <- data %>% filter(Gender == 2)
female_data <- data %>% filter(Gender == 1)
male_sig=male_data[,c(4:8,127:131)] #select the significant brain regions
female_sig=female_data[,c(4:8,127:131)]


female_sig$Gender <- "Female"
male_sig$Gender <- "Male"

# Combine the datasets
combined_data <- rbind(female_sig, male_sig)

# Reshape the data to long format
library(tidyr)
long_data <- combined_data %>%
  pivot_longer(
    cols = starts_with("Trait") | starts_with("Region"),
    names_to = "Variable",
    values_to = "Value"
  )

# Calculate mean values for each group (Male/Female) and each variable
library(dplyr)
mean_data <- long_data %>%
  group_by(Gender, Variable) %>%
  summarise(Mean = mean(Value, na.rm = TRUE), .groups = "drop")

regions_data <- mean_data %>% filter(grepl("Region", Variable))

# Plot for Brain Regions
plot_regions_gender <- ggplot(regions_data, aes(x = Variable, y = Mean, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  labs(
    title = "Comparison of Mean Values by Gender: Brain Regions",
    x = "Brain Regions",
    y = "Mean Value",
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Male" = "lightblue", "Female" = "royalblue"))

##ggsave("brain_gender.png",plot=plot_regions_gender)
```

**VII. Re - Plot the heat map using the significant variables**

```{r}
sig_var=data[,c(1,2,4:8,127:131)]
correlation2=corrplot(cor(sig_var))
##ggsave("correlation2.png", plot = correlation2)
```
